### 2020-02-03 nlp_models.tex
updating nlp

Observe that these results indicate that $\alpha$ provides a good quality metric for comparing these two models, one of which which was designed to be of low quality.
The deficient GPT model has numerous weight matrices with unsually large fitted exponents (indicating that they are not HT at all).  
Indeed, we may expect that a poorly trained model will not exhibit consistent Heavy Tailed behavior in all layers.
On the other hand, as expected, the improved GPT2 model has, on average, smaller $\alpha$ than the older GPT, with all $\alpha\le6$.  
\michael{XXX.  FLESH OUT.}

@@

However, as seen in Figure \ref{fig:GPT-snorm-hist},
 the poorly trained GPT model has many smaller (log) Spectral Norms $\log\Vert\mathbf{W}\Vert_{\infty}$
than the GPT2, which would be inconsistent with a belief that smaller Spectral Norm is always better.
Indeed, because there are so many anamously small $\Vert\mathbf{W}\Vert_{\infty}$,
it appears that the GPT model may be exhibiting a kind of rank collapse.
This is an extremely important observation because it demonstrates that while the Spectral Norm
may correlate well with predicted test error, it is not a good indicator of the overall quality of a model,
and using it as an empirical metric may give spurious results when applied to poorly trained
or otherwise deficient models.

Note that Figure \ref{fig:GPT-snorm-hist} also shows a couple anomalously  large Spectral Norms.
From Figure \ref{fig:gpt-snorm-layer} (below), we see that these correspond to the first embedding layer(s).
These layers appear to have a normalization, and therefore a differenc scale.
For example, in GPT, most layers, the maximum eigenvalue $\lambda_{max}\sim\mathcal{O}(10-100)$,
but in the first embedding layer, the maximum is of order N (the number of words in the embedding), or
 $\lambda_{max}\sim\mathcal{O}(10^{5})$.  For GPT and GPT2, we layer all layers as-is (although one may to normalize
the first 2 layers by  $\mathbf{X}$ by $\frac{1}{N}$, or to treat them as an outlier).
Here, we do not include them in our computed average metrics in Table \ref{table:nlp},
and do not include them in the histogram plot in Figure \ref{fig:GPT-snorm-hist}.

>>>>

From this, we see that $\alpha$ provides a good quality metric for comparing these two models, the``bad'' GPT vs. the ``good'' GPT2.
The deficient GPT has numerous unsually large $\alpha$ exponents--meaning they are not Heavy Tailed at all.
Indeed, we expect that a poorly trained model lack Heavy Tailed behavior in all layers.
On the other hand, as expected, the improved GPT2 model has, on average, smaller $\alpha$ than the older GPT, with all $\alpha\le6$.

@@

However, as seen in Figure \ref{fig:GPT-snorm-hist},
the ``bad'' GPT model has, spuriously, many smaller Spectral Norms $\log\Vert\mathbf{W}\Vert_{\infty}$,
compared to the ``good'' GPT2--which violates the beleif that that smaller Spectral Norms are always better.
Indeed, because there are so many anonymously small $\Vert\mathbf{W}\Vert_{\infty}$,
it appears that the GPT model may be exhibiting a kind of soft rank collapse.
This is an extremely important observation because it demonstrates that while the Spectral Norm
may correlate well with predicted test error, it is not a good indicator of the overall quality of a model,
and using it as an empirical metric may give spurious results when applied to poorly trained
or otherwise deficient models.

Note that Figure \ref{fig:GPT-snorm-hist} also shows a couple unusually large Spectral Norms, but,
here, simply correspond to the change in normalization of the word embedding layer(s) and can be ignored.
From Figure \ref{fig:gpt-snorm-layer} (below), we see that these correspond to the first embedding layer(s).
These layers appear to have a different effectvie normalization, and therefore a different scale.
For example, in GPT, most layers, the maximum eigenvalue $\lambda_{max}\sim\mathcal{O}(10-100)$,
but in the first embedding layer, the maximum is of order N (the number of words in the embedding), or
 $\lambda_{max}\sim\mathcal{O}(10^{5})$.  For GPT and GPT2, we layer all layers as-is (although one may to normalize
the first 2 layers by  $\mathbf{X}$ by $\frac{1}{N}$, or to treat them as an outlier).
Here, we do not include them in our computed average metrics in Table \ref{table:nlp},
and do not include them in the histogram plot in Figure \ref{fig:GPT-snorm-hist}.
